---
title: Solutions for lab 4
date: 2026-09-17
author: George G Vega Yon
format: gfm
---

# Part 1

## Problem 1

```{r}
#| label: simulate_pi
simulate_pi <- function(n = 5e6) {
  inside <- 0
  for (i in 1:n) {
    x <- runif(1, -1, 1)
    y <- runif(1, -1, 1)
    if (x^2 + y^2 <= 1) inside <- inside + 1
  }
  4 * inside / n
}

set.seed(1)
simulate_pi(1e5) # Smaller value to render quicker
```

```{r}
#| eval: false
#| label: profvis
profvis::profvis({
  simulate_pi <- function(n = 5e6) {
    inside <- 0
    for (i in 1:n) {
      x <- runif(1, -1, 1)
      y <- runif(1, -1, 1)
      if (x^2 + y^2 <= 1) inside <- inside + 1
    }
    4 * inside / n
  }

  set.seed(1)
  simulate_pi()
})
```

The bottle neck is located in the `runif()`call. Calling this repeatedly is not ideal; instead, we can do a single call.

```{r}
#| label: simulate_pi2
simulate_pi2 <- function(n = 5e6) {

  points <- runif(n * 2, -1, 1) |> 
    matrix(ncol = 2)

  mean(rowSums(points^2) <= 1) * 4
}

microbenchmark::microbenchmark(
  simulate_pi(10000),
  simulate_pi2(10000),
  unit = "relative"
)
```

Notice this could be even faster if we skip the `matrix` step and instead use two calls to `runif()`--the matrix step has to allocate memory again.

## Problem 2

```{r}
#| label: make_kernel
#| cache: true
make_kernel <- function(coords, bandwidth) {
  d <- as.matrix(dist(coords))
  K <- exp(-(d / bandwidth)^2)
  K / rowSums(K)
}

simulate_spread <- function(coords, p0, nsteps = 300, bandwidth = 0.15) {
  p <- p0
  for (s in seq_len(nsteps)) {
    K <- make_kernel(coords, bandwidth)
    p <- as.vector(p %*% K)
    p <- p / sum(p)
  }
  p
}

set.seed(2)
k <- 1000
coords <- cbind(runif(k), runif(k))
p0 <- rep(1 / k, k)

p_final <- simulate_spread(coords, p0)
summary(p_final)
```

Let's create a plot to visualize the spatial distribution of the final probabilities.

```{r}
#| label: plot_spread
library(ggplot2)
df <- data.frame(x = coords[, 1], y = coords[, 2], p = p_final)
ggplot(df, aes(x = x, y = y, color = p)) +
  geom_point() +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Final Probability Distribution", color = "Probability")
```

Inspecting the profiling of the model, we can see that the bottleneck is in the `make_kernel()` function. Since the kernel does not require update, we can safely move it outside the loop:

```{r}
#| label: profvis_spread
#| cache: true
make_kernel <- function(coords, bandwidth) {
  d <- as.matrix(dist(coords))
  K <- exp(-(d / bandwidth)^2)
  K / rowSums(K)
}

simulate_spread2 <- function(coords, p0, nsteps = 300, bandwidth = 0.15) {
  p <- p0
  K <- make_kernel(coords, bandwidth) # Moving the kernel outside
  for (s in seq_len(nsteps)) {
    p <- as.vector(p %*% K)
    p <- p / sum(p)
  }
  p
}

microbenchmark::microbenchmark(
  simulate_spread(coords, p0, nsteps = 50),
  simulate_spread2(coords, p0, nsteps = 50),
  unit = "relative",
  times = 10
)
```

# Part 2


```{r}
#| label: flawed-poisson
#| eval: false
# Fit a Poisson regression (log link) by Newton-Raphson on the
# log-likelihood. See the lab for the score, Hessian, and the update
# equation this function is supposed to implement.

fit_poisson_nr <- function(X, y, maxit = 50, tol = 1e-8) {
  beta <- rep(0, ncol(X))
  beta[1] <- log(mean(y))

  for (it in 1:maxit) {
    mu <- exp(drop(X %*% beta))
    score <- crossprod(X, y - mu)
    info <- crossprod(X * mu, X)

    # Wrapping solve (which is when it fails)
    # in try catch so we can inspect it right there
    step <- tryCatch({
        solve(info, score)
    }, error = \(e) e)

    if (inherits(step, "error")) {
      browser() # Investigate the error in solve
      stop("There's a problem.")
    }

    beta <- beta - drop(step)

    if (max(abs(step)) < tol) break
  }

  list(coefficients = beta, iterations = it)
}

set.seed(331)
n <- 500
x1 <- rnorm(n, 50, 10)
x2 <- rbinom(n, 1, 0.4)
X <- cbind(1, x1, x2)
y <- rpois(n, exp(drop(X %*% c(-2, 0.05, 0.8))))

fit_poisson_nr(X, y)
```

This was a hard problem to identify. The issue was that we had the wrong sign in the update step. Since we are using the information matrix, we need to add it to beta, not substract it


```{r}
fit_poisson_nr_corrected <- function(X, y, maxit = 50, tol = 1e-8) {
  beta <- rep(0, ncol(X))
  beta[1] <- log(mean(y))

  for (it in 1:maxit) {
    mu <- exp(drop(X %*% beta))
    score <- crossprod(X, y - mu)
    info <- crossprod(X * mu, X)

    # Wrapping solve (which is when it fails)
    # in try catch so we can inspect it right there
    step <- tryCatch({
        solve(info, score)
    }, error = \(e) e)

    if (inherits(step, "error")) {
      browser() # Investigate the error in solve
      stop("There's a problem.")
    }

    # HERE WAS THE ISSUE
    # beta <- beta - drop(step)
    beta <- beta + drop(step)

    if (max(abs(step)) < tol) break
  }

  list(coefficients = beta, iterations = it)
}

set.seed(331)
n <- 500
x1 <- rnorm(n, 50, 10)
x2 <- rbinom(n, 1, 0.4)
X <- cbind(1, x1, x2)
y <- rpois(n, exp(drop(X %*% c(-2, 0.05, 0.8))))

fit_poisson_nr_corrected(X, y)
```