---
title: Solutions for lab 7
date: 2026-10-08
author: George G Vega Yon
format: gfm
---

## Question 1: Overhead cost

```{r}
set.seed(123)
x <- runif(n=100)

serial_sum <- function(x){
  x_sum <- sum(x)
  return(x_sum)
}
```


```{r}
library(parallel)
set.seed(123)
x <- runif(n=100)

parallel_sum <- function(x, n_cores = 4) {
  
  # Set number of cores to use
  cl <- parallel::makePSOCKcluster(n_cores)

  # Split the data into chunks for each core
  indices <- parallel::splitIndices(length(x), n_cores)
  x_split <- lapply(indices, \(i) x[i])
  
  # Calculate partial sums doing something like:
  partial_sums <- parallel::parSapply(cl, x_split, sum)
  
  # Stop the cluster
  parallel::stopCluster(cl)
  
  # Add and return the partial sums
  sum(partial_sums)
  
}
```

Since summing a "small" number of vectors in R is very fast, the overhead cost of parallelizing the work increases the compute time significantly; this is why parallelizing everything is not a good solution. Now, this type of problems (in which you have to split the data yourself) are not rare, so just knowing how to do this is useful. Here is the benchmark of the two functions:

```{r}
bench::mark(serial=serial_sum(x),parallel=parallel_sum(x),relative=TRUE)
```

Just for completeness, we are doing a different approach to parallelization: passing the indices to the parallel function instead of splitting the data before. 


```{r}
parallel_sum2 <- function(x, n_cores = 4) {
  
  # Set number of cores to use
  cl <- parallel::makePSOCKcluster(n_cores)

  # make cluster and export to the cluster the x variable
  # Use "split function to divide x up into as many chunks as the number of cores
  indices <- parallel::splitIndices(length(x), n_cores)
  parallel::clusterExport(cl, "x")
  
  # Calculate partial sums doing something like:
  partial_sums <- parallel::parSapply(cl, indices, \(i) sum(x[i]))
  
  # Stop the cluster
  parallel::stopCluster(cl)
  
  # Add and return the partial sums
  sum(partial_sums)
  
}

bench::mark(
  serial=serial_sum(x),
  parallel=parallel_sum(x),
  parallel2=parallel_sum2(x),
  relative=TRUE
  )
```

Since we added the overhead of exporting the variable to the cluster, the second parallelization is even slower than the first one.

## Question 2: Power calculation


```{r}
library(parallel)

# 1. A function that simulates ONE dataset of size n using a given seed,
#    fits the probit model, and returns TRUE if H0 is rejected
sim_one <- function(n, seed, beta = c(0, 0.5, 1), alpha = 0.05) {

  set.seed(seed)

  # Simulate x, d (exactly n/2 treated), and y
  x <- rnorm(n)
  d <- rep(c(0, 1), each = n/2)
  y <- rbinom(n, 1, prob = pnorm(beta[1] + beta[2] * x + beta[3] * d))

  # Fit the probit model and extract the z-value for d
  summ_table <- glm(y ~ x + d, family = binomial(link = "probit")) |>
    summary() |>
    coef()
  
  # Return TRUE/FALSE
  summ_table["d", "Pr(>|z|)"] < alpha

}
```

```{r}
#| eval: false
# 2. The simulation grid: one row per (sample size, replicate), with its
#    own seed drawn ONCE in the main session
set.seed(7045)
sizes <- seq(100, 1000, by = 100)
nsim  <- 500
grid  <- expand.grid(rep = 1:nsim, n = sizes)
grid$seed <- sample.int(.Machine$integer.max, nrow(grid))

# 3. Run the simulation in serial
ans_serial <- lapply(seq_len(nrow(grid)), function(i) {
  sim_one(grid$n[i], grid$seed[i])
})

# 4. Run the simulation in parallel (create the cluster, export what
#    the workers need, run parLapply, stop the cluster)

# 5. Check that both are the same and compute the power by sample size
identical(ans_serial, ans_parallel)
tapply(unlist(ans_parallel), grid$n, mean)
```